---
title: "Anlysis   the   relationship   between   influenza   vaccination   coverage   and   Pneumonia   and   Influenza (P&I)   mortality   rates   by   state   in   the   US (2009-2019)."
author: "Wenjia Lu"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
---

This is my PM566 Final Project website.

<br>

#### *Link to my written report:      [[Download]](https://github.com/WennyLuu/PM566_Final/blob/main/PM566_Final_Report.pdf)*

<br>

## **Research Question**

What is the relationship between influenza vaccination coverage and Pneumonia and Influenza (P&I) mortality rates by state in the United States during 2009 to 2019? 

<br>

## **Why Study this?**

The objective of this study is to examine the relationship between influenza vaccination coverage and P&I mortality rates during the specified time frame. Understanding this relationship can provide insights into the effectiveness of vaccination programs and the impact of vaccination on P&I-related mortality. It also allows for an assessment of geographic variations, highlighting areas with successful vaccination programs and areas where improvements may be needed.

<br>

## **Dataset Description**

There are two datasets included in this analysis:

- Pneumonia and Influenza (P&I) Deaths Data comes from National Center For Health Statistics Mortality Surveillance System (NCHS), a division of the Centers for Disease Control and Prevention (CDC).
    
- Influenza Vaccinations Coverage Data comes from the National Immunization Survey-Flu (NIS-Flu) and the Behavioral Risk Factor Surveillance System (BRFSS).

<br>

## **Results**

```{r, message=FALSE, echo=FALSE, warning=FALSE, include=FALSE}

library(data.table)
library(dplyr)
library(tidyverse)
library(leaflet)
library(dtplyr)
library(knitr)
library(ggplot2)
library(lubridate)
library(usmap)
library(plotly)
library(DT)
```

```{r, include=FALSE} 
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  eval=TRUE,
  echo = TRUE,
  cache = FALSE,
  fig.width = 7, 
  fig.align = 'center',
  fig.asp = 0.618,
  out.width = "700px",
options(repos = c(CRAN = "http://cran.rstudio.com")))
```

```{r, include=FALSE}
## Read in data
death_dat <- data.table::fread("~/Desktop/PM566/DATA/Deaths_from_P&I.csv")
vac_dat <- data.table::fread("~/Desktop/PM566/DATA/Vaccination.csv")

## Check the data 

```

```{r, include=FALSE}
## Rename some variables
death_dat <- as_tibble(death_dat)
vac_dat <- as_tibble(vac_dat)

death_dat <- death_dat[which(death_dat$geoid == "State"), c(3, 5, 9, 11)]
names(death_dat) <- c("state", "season", "deaths_num", "deaths_prop" )

vac_dat <- vac_dat[which(vac_dat$'Geography Type' == "States/Local Areas" & vac_dat$'Dimension Type' == "Age"), c(3, 5, 8, 9, 11)]
names(vac_dat) <- c("state", "season", "age_group", "vac_prop", "sample_size" )
vac_dat <- vac_dat[!(vac_dat$season %in% c("2019-20","2020-21", "2021-22", "2022-23")), ]


## combine age groups: 0-18, 18-65 and 65+
vac_dat <- as.data.table(vac_dat)
vac_dat[, age_group := fifelse(
  age_group %in% c("6 Months - 17 Years","≥6 Months", "13-17 Years","5-12 Years", 
                   "6 Months - 4 Years"), "0~18 Years",
  fifelse(age_group == "≥65 Years" , "65+ Years",
          fifelse(age_group %in% c("18-49 Years at High Risk", "18-49 Years", 
                                   "18-64 Years not at High Risk", "18-64 Years",
                                   "18-64 Years at High Risk", "18-49 Years not at High Risk","≥18 Years", "50-64 Years", "25-64 Years not in Initial Target Group", "25-64 Years at High Risk", "6 Months - 64 Years at High Risk (Initial Target Group)"), "18~65 Years", age_group)
          )
  )
  ]

## Impute missing values as mean or mode
death_dat$deaths_num[is.na(death_dat$deaths_num)] <- mean(death_dat$deaths_num, na.rm=TRUE)
death_dat$deaths_prop[is.na(death_dat$deaths_prop)] <- mean(death_dat$deaths_prop, na.rm=TRUE)

vac_dat <- vac_dat[!(vac_dat$vac_prop %in% c("NR †", "NR *", "NR")), ]
vac_dat$vac_prop <- as.numeric(vac_dat$vac_prop)
vac_dat$vac_prop[is.na(vac_dat$vac_prop)] <- mean(vac_dat$vac_prop, na.rm=TRUE)

## Caculate some values
vac_dat$sample_size <- as.numeric(vac_dat$sample_size)
vac_dat$vac_num <- round(vac_dat$vac_prop / 100 * vac_dat$sample_size)

```

```{r, include=FALSE}
## Wrangle the data
vac_dat <- vac_dat %>% 
  group_by(state, season, age_group) %>% 
  mutate(
    vac_prop = mean(vac_prop),
    sample_size = mean(sample_size),
    vac_num = round(vac_num, 1)
  )
vac_dat <- distinct(vac_dat)

death_dat <- death_dat %>% 
  group_by(state, season) %>% 
  mutate(
    deaths_prop = round(mean(deaths_prop), 2),
    deaths_num = round(sum(deaths_num), 1)
  )
death_dat <- distinct(death_dat)

combined_dat <- merge(vac_dat, death_dat, 
                       by = c("state", "season"), 
                       all.x = TRUE, all.y = FALSE
                      )

combined_dat$deaths_num[is.na(combined_dat$deaths_num)] <- mean(combined_dat$deaths_num, na.rm=TRUE)
combined_dat$deaths_prop[is.na(combined_dat$deaths_prop)] <- mean(combined_dat$deaths_prop, na.rm=TRUE)

```

<br>

### 1. Regional variations in P&I (Pneumonia and Influenza) death rates and flu vaccination rates

## {.tabset}

### Vaccination Proportion

```{r, echo=FALSE}
set_map_details <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('white')
)

vac_map <- combined_dat %>% select(state, vac_prop)

st_crosswalk <- tibble(state = state.name) %>%
  bind_cols(tibble(abb = state.abb))
vac_map2 <- left_join(vac_map, st_crosswalk, by="state")
vac_map2$state.name <- vac_map2$state
vac_map2$state <- vac_map2$abb
vac_map2$abb <- NULL
vac_map2$hover <- with(vac_map, paste(state.name, '<br>', "Vaccination Proportion (%):", vac_prop))

f6 <- plot_geo(vac_map2, locationmode = 'USA-states') %>%
  add_trace(
    z = ~vac_prop,
    text = ~hover,
    locations = ~state,
    color = ~vac_prop,
    colors = 'Reds'
  )

f6 <- f6 %>% colorbar(title = "Vaccination Proportion (%)")
f6 <- f6 %>% layout(
  title = "Fully Flu Vaccination Proportion in Different States in the United States",
  geo = set_map_details
)

f6
```

### Death Proportion

```{r, out.width ="75%", out.height= "75%", echo=FALSE}
death_map <- combined_dat %>% select(state, deaths_prop)

st_crosswalk <- tibble(state = state.name) %>%
  bind_cols(tibble(abb = state.abb))
death_map2 <- left_join(death_map, st_crosswalk, by="state")
death_map2$state.name <- death_map2$state
death_map2$state <- death_map2$abb
death_map2$abb <- NULL
death_map2$hover <- with(death_map, paste(state.name, '<br>', "Death Proportion:", deaths_prop))

f5 <- plot_geo(death_map2, locationmode = 'USA-states') %>%
  add_trace(
    z = ~deaths_prop,
    text = ~hover,
    locations = ~state,
    color = ~deaths_prop,
    colors = 'Blues'
  )

f5 <- f5 %>% colorbar(title = "Deaths Proportion (%)")
f5 <- f5 %>% layout(
  title = "P&I Deaths Proportion in Different States in the United States",
  geo = set_map_details
)

f5
```

## {-}

* Hawaii and Louisiana have the highest and lowest P&I deaths rates. PA-Rest of state and Puerto Rico have the highest and lowest proportion of flu vaccination rates. 
* Hawaii, North Dakota and California have the highest P&I deaths proportion. At the same time, Louisiana, Florida and Oregon have the lowest P&I deaths proportion.

<br>

### 2. Yearly Fluctuations of P&I (Pneumonia and Influenza) death rates and flu vaccination rates

```{r, echo=FALSE}
f1 <- ggplot(combined_dat, aes(x = season)) +
  geom_point(aes(y = vac_prop, color = "Vaccination Proportion (%)"), size = 2) +
  scale_color_manual(
    values = c("Vaccination Proportion (%)" = "red"),
    name = "") +
  ggtitle("Fully Flu Vaccination Proportion by Year(2009-2019)") +
  labs(
    x = "Year",
    y = "Proportion (%)",
    color = "Legend"
  ) + 
  theme(legend.position = "top") +
  scale_y_continuous(breaks = seq(0, 70, by = 10))

f2 <- ggplot(combined_dat, aes(x = season)) +
  geom_point(aes(y = deaths_prop, color = "Deaths Proportion (%)"), size = 2) +
  scale_color_manual(
    values = c("Deaths Proportion (%)" = "blue"),
    name = "") +
  ggtitle("P&I Deaths Proportion by Year(2009-2019)") +
  labs(
    x = "Year",
    y = "Proportion (%)",
    color = "Legend"
  ) + 
  theme(legend.position = "top") +
  scale_y_continuous(breaks = seq(0, 70, by = 10))

```

## {.tabset}

### Vaccination Proportion
```{r echo=FALSE}
ggplotly(f1)
```

### Death Proportion
```{r echo=FALSE}
ggplotly(f2) 
```

## {-}

* The mortality rate show some variability across the years from 2009 to 2019, and the vaccination rate changed slightly, especially from 2017 to 2019 when it increased significantly.

<br>

### 3. The Relationship between P&I (Pneumonia and Influenza) death rates and flu vaccination rates

```{r, out.width ="75%", out.height= "75%", echo=FALSE}
f3 <- combined_dat %>% 
  ggplot(mapping = aes(x = vac_prop, y = deaths_prop)) + 
  geom_point(mapping = aes(color = state)) + 
  geom_smooth(method = lm, se = FALSE) + 
  labs(title = "Fully Vaccination Proportion vs P&I Deaths Proportion (All States)", 
       x  = "Vaccination proportions (%)", y = "P&I Deaths Proportion proportions (%)") +
  theme(legend.key.width = unit(0.01, "cm"),  
        legend.key.height = unit(0.1, "cm"),
        legend.position = "right") +
  guides(color = guide_legend(ncol = 2, title = NULL))

```

## {.tabset}

### Correlation by State

```{r echo=FALSE}
ggplotly(f3) 
```

## {-}

* There is a significant positive relationship between fully Vaccination proportion and P&I deaths proportion, indicating that higher vaccination coverage tends to be associated with higher mortality rates due to pneumonia and influenza.

```{r, out.width ="75%", out.height= "75%", echo=FALSE, warning=FALSE}
cor <- combined_dat %>%
  group_by(state) %>%
  summarize(cor = cor(deaths_prop, vac_prop)) %>%
  filter(!is.na(cor)) %>%
  mutate(cor = round(cor, 3)) %>%
  arrange(desc(cor))

```

## {.tabset}

### Correlation Table

```{r echo=FALSE}
datatable(cor, rownames = F) 
```

## {-}

* Utah has the strongest positively correlation bewteen flu vaccination rates and P&I deaths rates and District of Columbia	has the strongest negatively correlation.


## {-}
## **Conclusion**

In summary, the data suggests that there are significant state variations in P&I deaths and flu vaccination rates in the U.S. with year-to-year fluctuations. While vaccination rates have generally improved in recent years, there are some exceptions. The correlation analysis highlights that vaccination efforts can have a positive impact on reducing P&I deaths in some states. However, specific factors contributing to these trends would require further investigation.
